Gradual Soundness: Lessons from Static Python

### 2022-01-16 pj-7.1/paper.tex
pj: reword CPU utilization

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested using
  property-based soundness tests and \numSPtest{} tests from the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are based on production experience in the Instagram
  web server.
  Converting modules to \SP{} over the past year has led to a \CPUchange{}
  improvement in CPU usage in the web server.

@@

Over the past year, Instagram has been applying gradual soundness to its
primary web server monolith.
The results are very positive.
Although only a handful of modules use \SP{} types (hundreds among thousands),
and only a few critical modules rely on concrete types and primitive types (dozens),
web server CPU usage has improved by \CPUchange{} according to internal profiling tools.
Type-directed optimizations outweigh the cost of enforcing soundness.
Consequently, we believe the details of the \SP{} approach are of interest to
the gradual typing community at large.

@@

%% 2022-01-05 Carl:
%% "All the data I’ve been sharing is from a single codebase, the primary IG web
%% server monolith. The existence of JITted but not SP modules is mostly just due
%% to the fact that opting a function into the JIT is semantically transparent and
%% (unless it reveals a latent bug in the JIT) requires no code changes, so it’s
%% quite easy to do (the only tradeoff pushing us away from JITting everything is
%% just how much CPU we spend JIT-compiling and how much memory we spend storing
%% JIT-compiled code), whereas opting a module into SP requires more work so the
%% bottleneck is available developer time to do conversions."

@@

%% 2022-01-14 TODO re 3.7%
%% That percentage represents the aggregate CPU improvement observed from all of
%% our individual diffs that converted modules to Static Python over the past
%% year. In a sizable number of cases a conversion to Static Python did mean that
%% functions in that module started being JITted that previously were not JITted,
%% because for most of the half we had a policy in place that all Static Python
%% modules are always JITted. (We’ve now decoupled these things for future
%% conversions, so opting functions into the JIT is now orthogonal to opting them
%% into SP.)
%% 
%% So the 3.7% does not include any CPU wins that Cinder or the JIT provide in any
%% non-static modules, it only includes improvements observed from specifically
%% converting a module from nonstatic to static. But it does in some cases include
%% the effects of going from nonstatic + no-JIT to SP+JIT for a given module. In
%% other cases a conversion may have gone from nonstatic+JIT to SP+JIT.
%% Unfortunately we don’t have any way to go back and tease apart these cases for
%% each individual module conversion.
%% 
%% What we are doing now is an experiment where we turn off SP in a set of hosts
%% and observe them relative to the rest of prod, without changing the JIT list.
%% This will show some % regression, which we expect to be some number <3.7%, and
%% that will give a solid figure for the benefit we are getting specifically from
%% SP across all static modules, totally independent of the JIT. Unfortunately
%% there have been some hiccups in this experiment and we don’t have that number
%% ready yet, but we hope to have it in the next week or so.

%% 2022-01-14 FILL TODO 
%% > I still have trouble understanding the CPU number. There must be many
%% > servers running many workloads and some statistics to merge them. Do
%% > you have a blog post or something that explains the profiling in more
%% > detail?
%% 
%% I don’t know of anything reasonably current in terms of blog posts. For IG web
%% prod, there are not “many workloads” — there is no specialization in the IG web
%% tier. There are many thousands of hosts, all running the exact same workload —
%% all those hosts are handling an undifferentiated random selection from a single
%% giant pool of incoming IG requests. So for all intents and purposes we can
%% consider them identical. (There are in practice some differences in the hosts
%% themselves — e.g. different generations of CPU hardware — that we do have to
%% take into account when running the experiments below.)
%% 
%% The details of how we measure a CPU regression or improvement are interesting,
%% though. What we do is we take a set of hosts that we want to measure (typically
%% an experiment group and a control group, where the control group is identical
%% to the rest of production and the experiment group has some change made to its
%% code or configuration whose effect we want to measure), and we slowly increase
%% the traffic to those hosts until they are at max CPU utilization — they cannot
%% handle any more traffic without falling over. Then we measure the
%% requests-per-second handled by each set of hosts; if the experiment group is
%% able to handle 2% more requests per second than the control group, while both
%% are at max CPU utilization, this effect is considered a 2% global CPU win.
%% 
%% This metric has the advantage that it is highly holistic and results-oriented;
%% it is a measure of our actual capacity to serve more requests with fewer hosts,
%% and thus it can subsume and account for, and give apples-to-apples comparison
%% of, many different possible reasons for an increase or decrease in a host’s
%% efficiency in serving requests.
%% 
%% The disadvantage of the metric is that the raw number it measures (requests per
%% second) is also highly dependent on the product code and client behavior at
%% that moment in time, and is not stable over time, because the average cost in
%% CPU cycles of “one request” is not stable over time. If the behavior of the IG
%% iOS or Android app changes so that it is sending more heavy requests and fewer
%% lightweight requests, RPS across the fleet will naturally decrease even though
%% the efficiency of the servers has not changed. Therefore, this metric is really
%% good for saying “at X point in time, Y change has a Z% overall effect on our
%% CPU efficiency” but it cannot be used to compare raw numbers from two different
%% points in time; we cannot use it to say “between X1 point in time and X2 point
%% in time, RPS increased Z%, therefore our efficiency has increased by Z%.”
%% 
%% So the 3.7% number we gave you is an aggregate win from doing many such
%% individual experiments over the course of the past year, where each experiment
%% was measuring the effect of converting X module (or small set of modules) from
%% nonstatic to static (where in some cases it was nonstatic+JIT to SP+JIT, and in
%% other cases it was nonstatic noJIT to SP+JIT.) And the updated number we hope
%% to get you soon is an overall measure from a single experiment, today, of the
%% effect of disabling SP everywhere, while leaving the JIT list untouched. We
%% expect the latter number to be smaller than 3.7% because it will remove the JIT
%% factor entirely, but there’s also an unknown factor (which could net positive
%% or negative) of how the behavior of long-ago-converted SP modules has changed,
%% or the capabilities of SP has changed, since their conversion, which changes
%% the net win of SP on those modules since the time when we converted them.

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools, which continuously monitor
performance, attribute a \CPUchange{} improvement in web server CPU usage to \SP{} conversions.
\Cref{t:prod-stat} reports a few other production statistics.
Over 500 modules have been converted thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites and enabling the JIT on certain functions.

\begin{table}
  \caption{Overview: \SP{} in production}
  \label{t:prod-stat}
  \begin{tabular}{l@{~~}l@{\qquad}l@{~~}l}
    %% 541 modules out of ~45000 total
    8     & months in production   & 541   & typed, strict modules \\
    \CPUchange{} & web server CPU improvement &   9   & using concrete types \\
    0     & net regressions        &  12   & using primitive types \\
    %% all prims are int32 and int64 and cbool
    %% 873 modules use JIT for some functions (about to increase by OOM)
  \end{tabular}
\end{table}

\subsection{Migration Path}

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython had improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow (with a
few compiled modules spread across an interpreted codebase).
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% reduction in web server CPU usage.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

\subsection{Analyzing Code Changes}

@@

%% TODO section formatting looks good?
\subsubsection{Code Changes for Performance}

@@

\subsection{Microbenchmarks}
%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
%% https://docs.google.com/spreadsheets/d/1IO25bL_8LwDqx3ciQTj9dMetsFnZFdiQIyjFvgYWIpI/edit?usp=sharing

>>>>

  \paragraph{Grounding}
  % What argument, feasibility proof, artifacts,
  % or results and evaluation support this work?
  Our model of \SP{} is implemented in PLT Redex and tested using
  property-based soundness tests and \numSPtest{} tests from the \SP{} regression suite.
  This paper includes a small core of the model to convey the main ideas
  of the \SP{} approach and its soundness.
  Our performance claims are based on production experience in the Instagram
  web server.
  Migrations to \SP{} in the server have caused a \CPUchange{} increase in
  requests handled per second at max CPU utilization.

@@

Over the past year, Instagram has been applying gradual soundness to its
primary web server monolith.
The results are very positive.
Although only a handful of modules use \SP{} types (hundreds among thousands),
and only a few critical modules rely on concrete types and primitive types (dozens),
these migrations have resulted in a \CPUchange{} increase in production requests handled
per second for servers at max CPU utilization.
Type-directed optimizations outweigh the cost of enforcing soundness.
Consequently, we believe the details of the \SP{} approach are of interest to
the gradual typing community at large.

@@

@@

@@

The Instagram web server has been running \SP{} in production since April 2021.
Overall, the results are very encouraging.
Instagram's internal profiling tools attribute a \CPUchange{} improvement
in CPU utilization~(\cref{s:what-is-cpu}) to \SP{} conversions~(\cref{s:migration-path}).
This is a big improvement at Instagram scale.

Developers have converted 541 modules to \SP{} thus far.
Despite some initial regressions, none of the converted modules
ran slower after small rewrites~(\cref{s:code-changes})
and/or enabling the JIT on certain functions.
Only 9 modules use concrete types and only 12 modules use primitive types,
but these features delivered critical performance in key modules.

%% all prims are int32 and int64 and cbool
%% 541 modules out of ~45000 total
%% 873 modules use JIT for some functions (about to increase by OOM)

\subsection{How to Interpret the CPU Utilization Result}
\label{s:what-is-cpu}

The Instagram web server code runs on a large number of host machines, each of
which continuously handles requests from a common pool.
Improvements to the server codebase should make these machines more efficient
as they handle arbitrary requests.

To measure the effect of a code change on these hosts, profiling tools select
two representative sets of machines to run as experimental and control groups.
The experimental groups gets the latest version of the server code;
the control group gets the previous version.
Next, the profiling tools slowly increase the number of requests that these
machines receive until each is running at max CPU utilization.
Once the machines are fully allocated, the tools measure how many requests
per second each group is able to handle.
If the experimental group can serve more requests per second, then the change
is recorded as a success.
Assuming that the experimental and control groups have equivalent hardware
capabilities, and assuming that both groups receive a representative sample
of requests, then the results of an experiment (increase/decrease in CPU utilization)
should predict actual performance.

The main threat to validity is that this CPU utilization number depends on
product code and on client behavior.
Over time, the size of the average request may increase and thereby increase
the CPU cost of serving one request.
Nevertheless, CPU utilization is the primary metric that Instagram
uses for production changes.

\subsection{Migration Path}
\label{s:migration-path}

\SP{} first entered the web server codebase as a replacement for a
few critical Cython modules.
Cython improved performance by compiling these modules to C,
but its partial adoption had led to an awkward workflow (with a
few compiled modules spread across an interpreted codebase).
Replacing these modules with \SP{} let programmers return to a conventional
Python workflow.
Furthermore, \SP{} types combined with Cinder resulted in a
0.7\% reduction in web server CPU usage.
Primitive types~(\cref{s:c-types}) were essential for matching Cython.

@@

\subsection{Analyzing Code Changes}
\label{s:code-changes}

@@

\subsubsection{Code Changes for Performance}

@@

\subsection{Microbenchmarks}
\label{s:microbenchmarks}
%% https://github.com/facebookincubator/cinder/tree/cinder/3.8/Tools/benchmarks
%% https://docs.google.com/spreadsheets/d/1IO25bL_8LwDqx3ciQTj9dMetsFnZFdiQIyjFvgYWIpI/edit?usp=sharing
